---
title: "Reproducible Research Assignment 1"
author: "JSS"
date: "July 26, 2016"
output: 
  html_document: 
    fig_caption: yes
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



## Goals for This Assignment
1. Load and preprocess the data
2. Make a histogram of the total number of steps taken each day and calculate and report the mean and median of the total number of steps taken per day
3. Make a time series plot to determine Which 5-minute interval, on average across all the days, contains the maximum number of steps
4. Calculate and report the total number of missing values in the dataset. Devise a strategy for filling in all of the missing values in the dataset.
5. Create a new dataset with the missing data filled in. Make a histogram of the total number of steps taken each day and Calculate and report the mean and     median total number of steps taken per day.
6. Look for pattern differences in weekends vs weekdays.

### Step 1: Load and preprocess the data
```{r}
activity_raw <- read.csv("activity.csv")
# Transform the date attribute to an actual date format
activity_raw$date <- as.POSIXct(activity_raw$date, format="%Y-%m-%d")

# Compute the weekdays from the date attribute
activity_raw <- data.frame(date=activity_raw$date, 
                           weekday=tolower(weekdays(activity_raw$date)), 
                           steps=activity_raw$steps, 
                           interval=activity_raw$interval)

# Compute the day type (weekend or weekday)
activity_raw <- cbind(activity_raw, 
                      daytype=ifelse(activity_raw$weekday == "saturday" | 
                                     activity_raw$weekday == "sunday", "weekend", 
                                     "weekday"))

# Create the final data.frame
activity <- data.frame(date=activity_raw$date, 
                       weekday=activity_raw$weekday, 
                       daytype=activity_raw$daytype, 
                       interval=activity_raw$interval,
                       steps=activity_raw$steps)

# Clear the workspace
rm(activity_raw)
```
### Step 2: Make Historgram and Report Mean and Median of Total Steps per Day
```{r}
# Compute the total number of steps each day (NA values removed)
sum_data <- aggregate(activity$steps, by=list(activity$date), FUN=sum, na.rm=TRUE)

# Rename the attributes
names(sum_data) <- c("date", "total")

# Compute the histogram of the total number of steps each day
hist(sum_data$total, 
     breaks=seq(from=0, to=25000, by=2500),
     col="blue", 
     xlab="Total number of steps", 
     ylim=c(0, 20), 
     main="Histogram of the total number of steps taken each day\n(NA removed)")
mean(sum_data$total)
median(sum_data$total)
```
The mean number of steps per day is **9354**.  The median number of steps per day is **10395**.

### Step 3: Make Time Series Plot to Determine Interval with Max Steps
```{r}
# Clear the workspace
rm(sum_data)

# Compute the means of steps accross all days for each interval
mean_data <- aggregate(activity$steps, 
                       by=list(activity$interval), 
                       FUN=mean, 
                       na.rm=TRUE)

# Rename the attributes
names(mean_data) <- c("interval", "mean")
# Compute the time series plot
plot(mean_data$interval, 
     mean_data$mean, 
     type="l", 
     col="blue", 
     lwd=2, 
     xlab="Interval [minutes]", 
     ylab="Average number of steps", 
     main="Time-series of the average number of steps per intervals\n(NA removed)")

# We find the position of the maximum mean
max_pos <- which(mean_data$mean == max(mean_data$mean))

# We lookup the value of interval at this position
max_interval <- mean_data[max_pos, 1]

# Clear the workspace
rm(max_pos, mean_data)

print(max_interval)
```
The internal with the maximum number of steps is interval **835**.

### Step 4: Report Number of Missing Values
```{r}
# Clear the workspace
rm(max_interval)

# A TRUE boolean value is equivalent to 1 and a FALSE to 0.
NA_count <- sum(is.na(activity$steps))
```
A the Dataset contains 2304 missing values.
To more accurately evalute, the missing values are replaced with the mean of the interval
```{r}
# Clear the workspace
rm(NA_count)

# Find the NA positions
na_pos <- which(is.na(activity$steps))

# Create a vector of means
mean_vec <- rep(mean(activity$steps, na.rm=TRUE), times=length(na_pos))
# Replace the NAs by the means
activity[na_pos, "steps"] <- mean_vec

# Clear the workspace
rm(mean_vec, na_pos)

```
### Make a Histogram that includes Missing Values Filled-In.  Report the calculated Mean and Median 
```{r}
# Compute the total number of steps each day (NA values removed)
sum_data <- aggregate(activity$steps, by=list(activity$date), FUN=sum)

# Rename the attributes
names(sum_data) <- c("date", "total")

# Compute the histogram of the total number of steps each day
hist(sum_data$total, 
     breaks=seq(from=0, to=25000, by=2500),
     col="blue", 
     xlab="Total number of steps", 
     ylim=c(0, 30), 
     main="Histogram of the total number of steps taken each day\n(NA replaced by mean value)")
# Calculate Mean and Median Values
mean(sum_data$total)
median(sum_data$total)
```

The mean number of steps per day is now calculated as **10766**.
The median number of steps per day is **10766**

### Is there a pattern of difference between steps take on a Weekday Versus a Weekend?

```{r}
# Clear the workspace
rm(sum_data)

# Load the lattice graphical library
library(lattice)

# Compute the average number of steps taken, averaged across all daytype variable
mean_data <- aggregate(activity$steps, 
                       by=list(activity$daytype, 
                               activity$weekday, activity$interval), mean)

# Rename the attributes
names(mean_data) <- c("daytype", "weekday", "interval", "mean")
# Compute the time serie plot
xyplot(mean ~ interval | daytype, mean_data, 
       type="l", 
       lwd=1, 
       xlab="Interval", 
       ylab="Number of steps", 
       layout=c(1,2))

# Clear the workspace
        rm(mean_data)
```
The plots show that most activity happens early in the day on Weekdays and that there is a generally higher level of activity throughout the day on Weekends
